---
title: "Getting started with spagam"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with spagam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Why spatial correction?

Plant breeding and genetics experiments are conducted across a range of
settings, from field plots and polytunnels to glasshouse benches and growth
chambers. Regardless of the setting, spatial variation is almost always present.
In field trials this might be soil fertility gradients, drainage patterns, or
edge effects. On glasshouse benches it could be light gradients, temperature
differences near heating pipes, or watering inconsistencies. This environmental
heterogeneity has nothing to do with genotype performance, but if it is not
accounted for, genotype rankings can be misleading and selection decisions
suffer.

Spatial correction is especially important when phenotypic data are used as
input to genome-wide association studies (GWAS). Spatially confounded phenotypes
inflate residual variance and reduce the power to detect true marker-trait
associations. By removing the spatial trend before GWAS, you obtain cleaner
genotype estimates (BLUEs or BLUPs) that give more reliable association signals
and better control of false positives.

**spagam** fits GAM-based spatial models (via `mgcv`) to separate the spatial
trend from the genotype signal. It supports:

- **Single-unit designs** (replicated or unreplicated) with BLUEs and/or BLUPs
- **Multi-unit (joint) designs** where a single model is fitted across several
  spatial units (e.g. multiple glasshouse benches or field blocks)

Throughout this vignette we use the term "bench" to refer to any discrete
spatial unit in an experiment, whether that is a glasshouse bench, a field
plot, a polytunnel bay, or a growth chamber shelf. In single-unit analyses the
bench concept does not apply; in multi-unit analyses each bench is modelled
with its own spatial surface while genotype effects are estimated jointly.

The package writes corrected genotype estimates, spatial trend surfaces,
diagnostic plots, and model summaries to an output directory so that results are
ready for downstream analysis or archiving.

## Installation

```{r install, eval = FALSE}
# Install from GitHub
# install.packages("remotes")
remotes::install_github("alexd106/spagam")
```

```{r load}
library(spagam)
```

## Input data format

`correct_spatial()` expects a data frame with, at minimum:

| Column | Description |
|--------|-------------|
| **genotype** | Genotype or entry identifier (character or factor) |
| **row** | Row coordinate (numeric integer) |
| **col** | Column coordinate (numeric integer) |
| **phenotype(s)** | One or more numeric trait columns |
| **bench** *(optional)* | Identifier for the spatial unit (bench, block, etc.) |

Column names are flexible and you specify them via function arguments.
Coordinates should be integer grid positions (1, 2, 3, ...), not metres or GPS
coordinates.

For this vignette we will use two datasets:

1. The bundled `wheatdata`, a replicated wheat yield trial (single unit)
2. A simulated multi-bench dataset generated with `simulate_field_trial()`

### Wheat data (replicated, single unit)

```{r wheat-data}
data(wheatdata)
str(wheatdata)
```

### Simulated data (unreplicated, multi-bench)

```{r sim-data}
sim <- simulate_field_trial(
  n_bench           = 2,
  n_rows            = 10,
  n_cols            = 20,
  spatial_type      = 5,
  spatial_intensity = 25,
  save_csv          = FALSE,
  seed              = 42
)
str(sim$data)
```

We can visualise the raw spatial pattern across benches:

```{r sim-heatmaps, fig.width = 10, fig.height = 4}
plot_bench_heatmap(sim$data, pheno = "BNI", bench_col = "Bench",
                   bench = "Bench1", bench_label = "Bench 1")
```

## Example 1: Replicated single-unit design

The `wheatdata` trial has 107 genotypes across 3 replicates in a 22 x 15 grid.
We fit a spatial model and extract BLUEs (best linear unbiased estimates):

```{r single-bench, results = "hide", message = FALSE, warning = FALSE}
out_single <- correct_spatial(
  data          = wheatdata,
  pheno_cols    = "yield",
  geno_col      = "geno",
  row_col       = "row",
  col_col       = "col",
  estimate_type = "BLUEs",
  output_dir    = file.path(tempdir(), "wheat_output")
)
```

### What comes back?

`correct_spatial()` returns a list (invisibly) and writes files to `output_dir`.

```{r single-result}
names(out_single)
head(out_single$blues)
head(out_single$model_summary)
```

### Output files

The output directory now contains:

```{r single-files}
list.files(file.path(tempdir(), "wheat_output"))
```

| File | Contents |
|------|----------|
| `BLUEs.csv` | Corrected genotype estimates and standard errors |
| `spatial_trends.csv` | Per-plot spatial smooth and total spatial effect |
| `model_summary.csv` | Residual SD, effective degrees of freedom, convergence |
| `diagnostics_*.png` | 6-panel diagnostic plot (observed, spatial trend, residuals, variograms, QQ) |
| `row_col_re_*.png` | Row and column random effect magnitudes |
| `blues_distribution.png` | Histogram of corrected estimates |
| `spatial_surfaces.png` | Heatmap of the fitted spatial smooth |

### Diagnostic plots

The 6-panel diagnostic plot is the primary tool for checking model adequacy:

- **Top row:** Observed values | Fitted spatial trend | Residuals
- **Bottom row:** Variogram of raw data | Variogram of residuals | QQ plot

A well-fitting model should show a flat residual variogram (spatial
autocorrelation removed) and residuals that follow the QQ line.

## Example 2: Unreplicated multi-bench design

For multi-bench experiments, provide the `bench_col` argument. The package fits
a joint GAM with bench-specific spatial surfaces:

```{r multi-bench, results = "hide", message = FALSE, warning = FALSE}
out_multi <- correct_spatial(
  data          = sim$data,
  pheno_cols    = "BNI",
  geno_col      = "Genotype",
  row_col       = "Row",
  col_col       = "Col",
  bench_col     = "Bench",
  estimate_type = "BLUEs",
  output_dir    = file.path(tempdir(), "sim_output")
)
```

### Results

```{r multi-result}
head(out_multi$blues)
out_multi$model_summary
```

The model summary now has one row per bench, so you can compare residual
variation and effective degrees of freedom across benches.

### Output files

```{r multi-files}
list.files(file.path(tempdir(), "sim_output"))
```

Diagnostic and random effect plots are produced per bench, so you can assess
model fit for each bench independently.

## BLUPs (single-unit only)

For single-unit designs you can also extract BLUPs (best linear unbiased
predictors), where genotype is treated as a random effect. This shrinks
estimates towards the population mean and is useful when the number of
genotypes is large relative to replication. BLUPs are often preferred as input
to GWAS because the shrinkage reduces noise in the phenotypic values.

```{r blups, results = "hide", message = FALSE, warning = FALSE}
out_blups <- correct_spatial(
  data          = wheatdata,
  pheno_cols    = "yield",
  geno_col      = "geno",
  row_col       = "row",
  col_col       = "col",
  estimate_type = "both",
  output_dir    = file.path(tempdir(), "wheat_blups")
)
```

```{r blups-result}
head(out_blups$blups)
```

Note: BLUPs are not supported in multi-bench mode. Attempting
`estimate_type = "BLUPs"` with `bench_col` set will raise an informative error.

## Tips

- **Automatic basis dimensions:** By default, the number of B-spline knots is
  chosen adaptively (roughly half the unique positions per axis, bounded to
  5 to 20). Override with `k_row` / `k_col` if the spatial surface looks
  under-smoothed or over-smoothed.
- **Multiple traits:** Pass several column names to `pheno_cols` to correct
  multiple traits in one call. Each trait gets its own model, diagnostics, and
  output columns.
- **Reproducibility:** All output CSVs and plots are written to `output_dir`,
  making it straightforward to archive results alongside analysis scripts.
